---
title: "Statewide Analysis"
---

## Setup

```{r setup}
library(tidyverse)
library(janitor)
library(lubridate)
```


## Read in Clean Data

```{r read}
df <- read_rds("data-processed/01-storm-data.rds")

df |> glimpse()

```

# Answering Questions

## Which storm types cause the most damage?

```{r filter-property}
#filter for only events that have damage values for property
damages_property <- df |> filter((!is.na(damage_val_prop)))

damages_property
```

```{r most-damage-prop}
#now groupby event type and sum up damages
most_damage_property <- damages_property |> group_by(EVENT_TYPE) |> 
  summarize(total_damage_property = sum(damage_val_prop)) |> 
  arrange(total_damage_property |> desc())

most_damage_property
```

Flash Floods caused the most damage to property in Texas since 2000, totaling over $48 billion in damage. Hail was the second most damaging event type with $11.8 billion in damage. 

```{r filter-crop}
damages_crops <- df |> filter((!is.na(damage_val_crop)))

damages_crops
```


```{r most-damage-crop}
most_damage_crop <- damages_crops |> group_by(EVENT_TYPE) |> 
  summarize(total_damage_crop = sum(damage_val_crop)) |> 
  arrange(total_damage_crop |> desc())

most_damage_crop
```

Droughts caused the most damage to crops in Texas out of all severe weather events with $9.9 billion of damage.

## Where did the most damaging storms occur?

```{r sort-prop}
sorted_damages_prop <- damages_property |> arrange(damage_val_prop |> desc())

sorted_damages_prop
```

```{r sort-crop}
sorted_damages_crop <- damages_crops |> arrange(damage_val_crop |> desc())

sorted_damages_crop
```

## What year had the most damages?

```{r most-year}
yr_damage_prop <- damages_property |> group_by(year(begin_date)) |> 
  summarize(total_damage_prop = sum(damage_val_prop)) |> 
  arrange(total_damage_prop |> desc())

yr_damage_crop <- damages_crops |> group_by(year(begin_date)) |> 
  summarize(total_damage_crop = sum(damage_val_crop)) |> 
  arrange(total_damage_crop |> desc())

yr_damage_prop
yr_damage_crop
```

## How have the damages from storm events changed over time?

```{r plot-prop}
ggplot(
  damages_property,
  aes(x = begin_date, y = damage_val_prop)
) +
  geom_point()
```


```{r plot-crop}
ggplot(
  damages_crops,
  aes(x = begin_date, y = damage_val_crop)
) +
  geom_point()
```

