---
title: "Interactive Data Table using Reactable"
---

## Setup

```{r}
library(reactable)
library(tidyverse)
```

## Read in the data 

I am using the data from O2-BasicAnalysis because it has the correct columns already.

```{r}
storms <- read_rds("data-processed/02-storm-data.rds")
```


## Reactable Interactive Table

I wanted to create an interative data table that I could search and filter for records. To do this with reactable, I made the table "searchable" to add the top search bar. Then I made event type, location and county/zone filterable. I also made each damage column able to be arranges in desc or asc order with the defaultSortOrder. I did the same with the date column. 

```{r}
reactable(
    storms, 
    searchable = TRUE,
    columns = list(
    event_type = colDef(name = "Event Type", filterable = TRUE),
    location = colDef(name = "Location", filterable = TRUE),
    CZ_type = colDef(name = "County/Zone"),
    FIPS = colDef(name = "FIPS Code", filterable = TRUE),
    total_damages = colDef(name = "Total Damages", align = "center", defaultSortOrder = "desc"),
    damage_val_prop = colDef(align = "center", name = "Damages to Property", defaultSortOrder = "desc"),
    damage_val_crop = colDef(align = "center", name = "Damages to Crops", defaultSortOrder = "desc"),
    begin_date = colDef(name = "Start Date", defaultSortOrder = "asc")
    )
  )
```

